# The Uniform Distribution {#uniform_distribution}

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(mosaic)
set.seed(2018)
```

## Introduction {#introduction}

We have learned about important discrete random variables such as the binomial and about important continuous random variables such as the normal. In this chapter we add to our repertoire some other useful distributions including the geometric, negative binomial, and poisson discrete random variables and the uniform and exponential continuous random variables. 

## Chapter Scenario - Lovin' the Airport {#chapter_scenario_lovin_the_airport}

Suppose a domestic flight leaves at 5:00pm and has 50 passengers who show up independently at the airport at uniformly random times between 3:15pm and 4:30pm. How many passengers can we expect to have arrived at the airport by 4:00pm, one hour before takeoff? How confident can we be in this answer?

## The Uniform Distribution {#the_uniform_distribution}

The uniform random variable is a continuous random variable that is equally likely along its entire range of possible values. Like the normal distribution, the area under the curve between two values is the probability of the random variable being in the interval but unlike the normal bell-shaped curve, the uniform probability density function is a flat, horizontal line. If the uniform random variable X ranges from a minimum of $a$ to a maximum of $b$, we abbreviate this $X \sim UNIF(a,b)$.

### The Uniform Probability Density Function {#the_uniform_probability_density_function}

The uniform probability density function is flat and the area under the curve equals the probability. For X, a uniformly distributed random variable with a minimum of a and a maximum of b, $X \sim UNIF(a,b)$, so that the total area under the curve is 1, the height of the curve must be $1/(b-a)$ between a and b and 0 elsewhere. The code below creates a function to plot different uniform random variables

```{r}
uniform_Plot <- function(a, b){
  xvals <- data.frame(x = c(a, b)) #Range for x-values
  ggplot(data.frame(x = xvals),aes(x = x)) + xlim(c(a, b)) + 
    ylim(0, 1/(b - a)) +
    stat_function(fun=dunif, args=list(min=a, max=b), geom = "area", 
                  fill="green", alpha=0.35) + 
    stat_function(fun = dunif, args = list(min = a, max = b)) +
    labs(x="\n u", y="f(u) \n", 
         title=paste0("Uniform Distribution \n With Min = ", a, " & Max = ", b, " \n")) +
    theme(plot.title=element_text(hjust = 0.5),
          axis.title.x=element_text(face="bold", color="blue", size=12),
          axis.title.y=element_text(face="bold", color="blue", size=12)) +
    geom_vline(xintercept=a, linetype="dashed", color="red") +
    geom_vline(xintercept=b, linetype="dashed", color="red")
}
```

Source: http://dkmathstats.com/plotting-uniform-distributions-in-r-with-ggplot2/ 

Here is the plot of UNIF(4,12).

```{r}
uniform_Plot(4,12)
```

For uniform random variable $X \sim UNIF(a,b)$, the expectation is $E(X)=\frac{a+b}{2}$ and the variance is $Var(X)=\frac{(b-a)^{2}}{12}$ and standard deviation $SD(X)=\sqrt{\frac{(b-a)^{2}}{12}}$.

### The Uniform Random Variable in R

For X a uniform random variable we use the following. Since by default `min=0` and `max=1`, for X that is uniformly distributed between a and b, $X \sim UNIF(a,b)$, we must replace 0 and 1 by a and b, respectively.
 
To find the probability of X being less than q, $P(X < q)$, use `punif(q, min = 0, max = 1, lower.tail = TRUE)`.

To find the probability of X being greater than q, $P(X > q)$, use `punif(q, min = 0, max = 1, lower.tail = FALSE)` or `1-punif(q, min = 0, max = 1, lower.tail = TRUE)`.

To find the inverse probability, that is the value of x such that $P(X \leq x) = p$ use `qunif(p, min = 0, max = 1, lower.tail = TRUE)`.

To generate a random sample of size n from a uniform random variable use `runif(n, min = 0, max = 1)`.


### Example - The UNIF(0,1) Random Variable

The basis for most random number generator schemes is a uniform random variable X with a minimum of 0 and a maximum of 1, $X \sim UNIF(0,1)$. The probability density function is visualized below.

```{r}
uniform_Plot(0,1)
```

Because the default `min=0` and `max=1` we could leave these parameters out when using `punif()`, `qunif()`, or `runif()`. For example, we can use `runif(n)` to generate a sample of size n. Here is a sample of twenty random UNIF(0,1):

```{r}
runif(n=20)
```

Variations on UNIF(0,1) is how other random numbers are generated. Watch what happens when we multiply `runif(n=20)` by 4.

```{r}
4*runif(n=20)
```

In this case, we stretched the range so that the values are now UNIF(0,4).

Suppose in addition to stretching by a factor of 4 we subtract 2.

```{r}
4*runif(n=20) - 2
```

This generates random uniform numbers UNIF(-2,2).

And if we had added 2 instead:

```{r}
4*runif(n=20) + 2
```

Now we are generating random uniform numbers UNIF(2,6).

Of course, we could generate all of these directly using `runif(0,4)`, `runif(-2,2)`, and `runif(2,6)` but now we know a little more about how R does it under the hood.

### Example - Catch a Flight

Suppose that the time check in at the airport for a 5:00pm flight is uniformly distributed from 3:15pm to 4:30pm. If we let X represent the amount of time in minutes after 3:15pm that a person shows up then $X \sim UNIF(0,75)$.

To find the probability a person chosen at random shows up before 4:15pm:

```{r}
punif(60, min = 0, max = 75, lower.tail = TRUE)
```

To find the probability a person chosen at random shows up after 4:15pm we have options:

```{r}
1-punif(60, min = 0, max = 75, lower.tail = TRUE)
punif(60, min = 0, max = 75, lower.tail = FALSE)
```

To find the 66th percentile, that is the amount of time after 3:15pm that $66\%$ of people show up:

```{r}
qunif(0.66, min = 0, max = 75, lower.tail = TRUE)
```

As a time, this would translate into 3:15pm plus `r qunif(0.66, min = 0, max = 75, lower.tail = TRUE)` minutes which would be 4:04:30pm.

To generate a random sample of 50 passengers and the number of minutes after 3:15pm that they show up:

```{r}
runif(n=50, min = 0, max = 75)
```

Doesn't this give you a feeling of power? It would be easy for this go to our head.

## Chapter Scenario Revisited - Lovin' the Airport {#chapter_scenario_revisited_lovin_the_airport}

Recall, a domestic flight leaves at 5:00pm, has 50 passengers who show up independently at uniformly random times between 3:15pm and 4:30pm and we want to know how many passengers we can expect to have arrived by 4:00pm and how confident can we be in this answer?

As we have seen, the time in minutes after 3:15pm that a passenger arrives can be modeled with a uniform random variable, $X \sim UNIF(0,75)$.

The probability an individual passenger chosen at random will have arrived by 4:00pm is

```{r}
punif(45, min = 0, max = 75, lower.tail = TRUE)
```

We have 50 passengers and the probability any one of them arrives by 4:00pm is `r punif(45, min = 0, max = 75, lower.tail = TRUE)`. The total number of passengers who have arrived by 4:00pm is thus like a binomial random variable with n=50 and p=`r punif(45, min = 0, max = 75, lower.tail = TRUE)`. The expected number of passengers to have arrived is `r 50*punif(45, min = 0, max = 75, lower.tail = TRUE)` but this could vary. Check out this distribution:

```{r}
xpbinom(20:40, size = 50, prob = 45/75)
```

While we expect around `r 50*punif(45, min = 0, max = 75, lower.tail = TRUE)` there is only an `r 100*dbinom(30, size=50, prob=45/75)` $\%$ chance of exactly `r 50*punif(45, min = 0, max = 75, lower.tail = TRUE)`. This kind of knowledge could help us know how to more effectively manage airport traffic.


## Exercises {#exercises}

### Exercise - Transforming UNIF(0,1)

(a) Describe a transformation on `runif(0,1)` that generates uniform random numbers with a min=0 and max=3.
(b) Describe a transformation on `runif(0,1)` that generates uniform random numbers with a min=-3 and max=3.
(c) Describe a transformation on `runif(0,1)` that generates uniform random numbers with a min=4 and max=12.

### Exercise - Showing Up for Class

Suppose that in a class of 20 students that begins at 2:00pm, the time each student shows up is independent of other students and is uniformly distributed from 1:55pm to 2:10pm.
(a) What is the probability a student chosen at random will arrive on time?
(b) At what time can we be 90% certain a student chosen at random has arrived?
(c) What is the probability student A shows up before student B?
(d) What is the probability that a majority of students will show up on time?

